---
title: "R Notebook"
output: html_notebook
---

Whenever you perform a one-way ANOVA, you will always compute three sum of squares values:

1. Treatment Sum of Squares (SST)

    This is the sum of the squared differences between each group mean and the grand mean.

2. Within Group Sum of Squares (SSE)

    This is the sum of the squared differences between each individual observation and the group mean of that observation.

3. Total Sum of Squares (TSS)

    This is the sum of the squared differences between each individual observation and the grand mean.

4. Sum of Squares Total (SST)
    SST = SST + SSE

```{r}
library(ggplot2)
library(dplyr)

iris %>% ggplot(aes(x = Species,y = Petal.Width, color = Species)) + geom_point() + theme_classic()
iris %>% ggplot(aes(x = Species,y = Petal.Length, color = Species)) + geom_point() + theme_classic()

```


```{r}

# Analysis of Variance calculation

sepal_aov <- aov(Sepal.Length ~ Species, data = iris)
summary(sepal_aov)
```


With a low p-value we reject the null hypothesis that there is no variance and accept the alternate hypothesis.

```{r}
petal_aov <- aov(Petal.Width ~ Species, data = iris)
summary(petal_aov)
```
For ANOVA testing the samples must be equal
Tukey Significance test: taking the ANOVA and analyzing each factor seperately

```{r}
TukeyHSD(sepal_aov)
```
```{r}
TukeyHSD(petal_aov)
```

Two-way ANOVA: Analyze 2 categorical variables against 1 continuous variable

```{r}
dataFileName <- "http://personality-project.org/r/datasets/R.appendix2.data"
data.ex2 <- read.table(dataFileName,header = T)
```

Analyzing weather dosage changes made a difference in alertness between men and women

```{r}
two_way <- aov(Alertness~ Gender* Dosage,data = data.ex2)
summary(two_way)
```
Cannot reject the null hypothesis that there is no variance with p value greater than 0.05

```{r}
boxplot(Alertness~ Gender + Dosage, data = data.ex2)
```



