---
title: "GAM_Brain_Activity"
author: "Thomas Wolff"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Bringing brain activity data into R Studio

```{r Brain Activity Data}
library(gamair)
library(mgcv)

data(brain)
```

##  Removing outliers from the medFPQ faeture and modeling with Gaussian model

The plots indicate the model residuals are not Gaussian and that some transformation needs to occur.

```{r Removing outliers, echo=TRUE}
brain <- brain[brain$medFPQ >= 3e-5,]

m0 <- gam(medFPQ~s(Y,X,k=100), data=brain)
gam.check(m0)
```
We can assume that the variance of y is proportional to mu raised to some power Beta 
where Beta is a parameter that we can estimate with regression

So we take the log of the squared residuals and regress using the fitted values

To find the value of Beta which is equal to 1.919 (2)

```{r Strip out residuals, echo=TRUE}
e <- residuals(m0); fv <- fitted(m0)
lm(log(e^2)~log(fv))

```
### Second model fit using a 4th-root transform on the response variable

This transform fixes a lot of the issues with the residuals and the fitted values

```{r Second Model Fit, echo=TRUE}
m1 <- gam(medFPQ^0.25~s(Y,X,k=100), data=brain)
gam.check(m1)

```



```{r Third Model Fit, echo=TRUE}
m2 <- gam(medFPQ~s(Y,X,k=100), data=brain, family=Gamma(link=log))
gam.check(m2)

mean(fitted(m1)^4)
mean(fitted(m2));(mean(brain$medFPQ))
```


### Plotting the brain activity using contour plot

This works well, but this model has many degrees of freedom, and is therefore too complex.

Using and additive model may work better.

```{r Contour Plot, echo=TRUE}

plot <- function(model) {
  vis.gam(model, plot.type = "contour", too.far=0.03,
          color="gray", n.grid=60, zlim=c(-1,2), 
          main=paste0("Model image plot and overlaid contours on scale of linear predictor"))
}

plot(m2)
```

### Here we try and additive model to simplify the model and reduce degrees of freedom

Where: log(E[medFPQi]) = f1(Yi) + f2(Xi) where medFPQ ~ Gamma

```{r Additive model Fit, echo=TRUE}
m3 <- gam(medFPQ~s(Y,k=30) + s(X, k=30), data=brain, family=Gamma(link=log))

gam.check(m3)

```
### New model with isotropic interaction term s(Y,X,k=100) added as a smoothing parameter.


```{r Additive model with isotropic interaction term, echo=TRUE}
m4 <- gam(medFPQ~s(Y,k=30) + s(X,k=30) + s(Y,X,k=100), data=brain, family=Gamma(link=log))

anova(m3,m4,test="F")

plot(m2)

plot(m3)

plot(m4)
```
### Tensor product smooths 

It is worth checking whether tensor product smooths are better than isotropic product smooths

Tensor products are scale invariant and are computationally efficient 

tm is a Tensor product of two cubic regression spline bases, rank 10 * rank 10 = rank 100

tm1 is a purely additive model with two rank 10 cubic regression splines so tm1 is nested within tm

```{r Tensor product model Fit, echo=TRUE}
## Tensor product on the interaction non-additive model
tm <- gam(medFPQ~te(Y,X,k=10), data=brain, family=Gamma(link=log))

tm1 <- gam(medFPQ~s(Y,k=10,bs="cr") + s(X,bs="cr",k=10), data=brain, family=Gamma(link=log))

```

