---
title: "Nonlinear Regression"
author: "Thomas Wolff"
date: "2024-05-23"
output: html_document
---

```{r setup, include=FALSE}

library(ISLR)
attach(Wage)

```

## R Markdown

Polynomial regression: Poly method creates orthogonal polynomials, regression with uncorrelated predictors

```{r cars}
fit <- lm(wage~poly(age,4),data=Wage)
summary(fit)
```

## Including Plots

You can also embed plots, for example:

```{r fig.width=7, fig.height=6}

agelims=range(age) # creating a grid of range age
age.grid=seq(from=agelims[1], to=agelims[2]) # creating a range of ages from beginning of age range to end (18 to 80)
preds=predict(fit,newdata=list(age=age.grid),se=TRUE)
# cbind: binding together columns to make a matrix
se.bands=cbind(preds$fit+2*preds$se,preds$fit-2*preds$se) # making +- 2 standard error bands with the fitted function
plot(age,wage,col="darkgrey") # plotting the data
lines(age.grid,preds$fit,lwd=2,col="blue") # plotting the fitted function
matlines(age.grid,se.bands,col="red",lty=2) # the standard error bands

```

Splines: more flexible than polynomials. Spline is a cubic polynomial (bs), with discontinuity in the 3rd derivative. They are constrained to be continuous at the knots (vertical dashed lines).

```{r}
library(splines)
fit <- lm(wage~bs(age,knots=c(25,40,60)),data=Wage)
```



```{r fig.width=7, fig.height=6}
plot(age,wage,col="darkgrey")
lines(age.grid,predict(fit,list(age=age.grid)),col="darkgreen",lwd=2) # fitting the model to data
abline(v=c(25,40,60),lty=2,col="darkgreen") # vertical lines at knots
```

```{r}
uswages = filter(uswages, wage<4000)
```


Polynomial Regression
```{r fig.width=9, fig.height=6}
library(splines)
library(faraway)
library(ggplot2)
library(dplyr)

g2 <- lm(wage~poly(exper, 2), data = uswages)
g4 <- lm(wage~poly(exper, 4), data = uswages)
uswagesPoly = mutate(uswages, degree2 = fitted(g2), degree4 = fitted(g4))
ggplot(uswagesPoly, aes(exper, wage)) +
  geom_point(col = "slategrey")+
  labs(x = "Experience in year", y = "Weekly wage")+
  geom_line(aes(exper, degree2, color = "2")) +
  geom_line(aes(exper, degree4, color = "4")) +
  scale_color_manual(values = c('2' = 'darkblue',
                                '4' = 'red')) +
  labs(color = 'Polynomial degree')+
  ggtitle("Polynomial Regression models")


```


Cubic Regression Splines

```{r}
quantiles <- quantile(uswages$exper,c(.25,.50,.75))
```

```{r fig.width=9, fig.height=6}

cubic_spline <- lm(wage~bs(exper, knots = quantiles), data=uswages)
uswages <- mutate(uswages, smooth = fitted(cubic_spline))
ggplot(uswages, aes(exper, wage)) +
  labs(x = "Experience in year", y = "Weekly wage") +
  geom_point(col = "slategrey") +
  geom_line(aes(exper, smooth), col = "red") +
  geom_vline(xintercept = quantiles, col = "green") +
  ggtitle("Cubic regression spline model: quantiles as knots")

```
```{r}
summary(cubic_spline)
```



```{r fig.width=9, fig.height=6}

knot_loc = c(15,30,45)

cubic_spline2 <- lm(wage~bs(exper, knots = knot_loc), data=uswages)
uswages <- mutate(uswages, smooth = fitted(cubic_spline2))
ggplot(uswages, aes(exper, wage)) +
  labs(x = "Experience in year", y = "Weekly wage") +
  geom_point(col = "slategrey") +
  geom_line(aes(exper, smooth), col = "red") +
  geom_vline(xintercept = knot_loc, col = "green") +
  ggtitle("Cubic regression spline model: evenly spaced knots")

```




```{r}
summary(cubic_spline2)
```


```{r}
library(knitr)

mse <- function(model) {
  rmse <- sqrt(mean(model$residual^2))
  round(rmse,3)
}

mse_poly1 <- mse(g2)
mse_poly2 <- mse(g4)
mse_cubic1 <- mse(cubic_spline)
mse_cubic2 <- mse(cubic_spline2)

model_name <- c("Polynomial with degree 2","Polynomial with degree 4","Cubic Spline Quartiles","Cubic Spline Intervals")
mses <- c(mse_poly1,mse_poly2,mse_cubic1,mse_cubic2)
mse_table <- cbind(model_name, mses)
colnames(mse_table) = c("Models","RMSE")
kable(mse_table, caption = "Mean squared errors for different models", digits = 3, table.attr = "class=\"striped\"", "html")

```




